Refinement of Thermostated Molecular Dynamics Using Backward Error Analysis

### 2019-02-20 response_to_reviewers.tex

\documentclass[]{article}
\usepackage{setspace}
\usepackage{xcolor}
\usepackage[version=4]{mhchem}
\usepackage{textcomp}
%opening
\title{\normalsize \textbf{Response to the Reviewers' Comments on Manuscript A18.12.0101}}
\author{}
\date{}

\begin{document}
\doublespacing
\maketitle

To: Prof. Carlos Vega

Associate Editor

The Journal of Chemical Physics
\vspace{1cm}

Dear Prof. Vega:

Thank you for your message forwarding the reviewer’s comments on our manuscript entitled “Refinement of Thermostated Molecular Dynamics Using Backward Error Analysis”. We have carefully addressed all comments, which required us to make changes in the text. In order to facilitate your evaluation, we have attached an annotated version in which we crossed out in red all removed text and incorporated new content in blue.

We hope the revised version of our manuscript is suitable for publication in the Journal of Chemical Physics.

\begin{center}
Sincerely yours,
\end{center}
\begin{flushright}
Ana J. Silveira

Memorial Sloan Kettering Cancer Center

\vspace{1cm}

Charlles R. A. Abreu

Federal University of Rio de Janeiro

\end{flushright}

\section{\textbf{Reviewer \#1}}

{\color{blue}{
The authors investigate the influence of finite time step numerical discretization errors on the computation of averages in the molecular system with rigid bodies. They analyze the possibility of compensating for such errors by using 'refined' expressions for the kinetic and potential energies and the re-weighting of computed quantities. While the effects of the numerical integration error and the re-weighting procedures have been studied before, the authors demonstrate how this approach can work in practice, using a model of rigid water molecules as an example.

Besides, the authors investigate the effect of using different thermostatting approaches, highlighting some of the pitfalls of using multiple thermostats, like in the KLN method. Unfortunately, the MD practitioners sometimes don't pay attention to such problems with their thermostats, which may lead to erroneous results.

Therefore, I recommend publication of this manuscript in the JCP, after the authors implement minor revisions listed below. 
}}

\textbf{Answer}:
We thank the reviewer for having done such a careful examination of our manuscript. All comments and suggestions were very meaningful and constructive. We really appreciate it.

{\color{blue}{
- Abstract: ``may distort such assumption". This needs re-wording, since equipartition is not an assumption. Maybe better say ``may lead to deviation from equipartition in measured quantities".
}}

\textbf{Answer}:
We replaced ``may distort such assumption'' by ``may lead to deviations from equipartition”. In order to maintain the specific reference to the kinetic energy, made in the preceding sentence, we did not include the fragment “in measured quantities".

{\color{blue}{
- Abstract: ``quantity that is actually subject to equipartition". I think this need to be stated more precisely, since we don't know the expression for such a quantity. What we can derive is a quantity that deviations from equipartition to higher order in the time step.
}}

\textbf{Answer}:
We replaced “identify” by “obtain a higher-order estimate of”.

{\color{blue}{
- Abstract: ``which coincides with'' better to say ``which converges to".
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.1, middle of right column: ``breakage'' \textrightarrow ``breakdown". 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.1, bottom of right column: ``manage to prove here'' - you are not proving, but rather ``show'' or ``demonstrate'' 
}}

\textbf{Answer}:
We were referring to the mathematical derivation in the Appendix. Anyways, we agree that this is not a formal proof and thus replaced “prove” by “demonstrate”.

{\color{blue}{
- p.2, middle of left column: ``specially'' \textrightarrow ``especially'' 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.2, bottom of right column: ``While U(r,q) is an arbitrary function". This may be interpreted too broadly. Better to say something like ``While the form of U(r,q) depends on the specific interaction model". 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.2, bottom of right column: ``all bodies piled together'' \textrightarrow ``all bodies combined in a single vector". 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.3, top of left column: ``freely moving atoms". I guess the authors talk about a mixed system of rigid bodies and point masses. Freely moving atoms has the meaning of 'non-interacting atoms'. I suggest replacing ``freely moving atoms'' with ``point masses'' or ``interacting point masses". 
}}

\textbf{Answer}:The reviewer is right. The expression ``freely moving atoms'' is not adequate. We decided to use the term “individual point masses” instead.

{\color{blue}{
- p.3, middle of left column: ``is a kick in the linear and quaternion momenta given by", better say ``is a kick that changes the linear and quaternion momenta according to'' 
}}

\textbf{Answer}:Done.

{\color{blue}{
- p.3, before subsection B: ``unplit'' \textrightarrow ``unsplit", although it may be better to say ``exact solution for free rotations'' instead. 
}}

\textbf{Answer}:
We used the second option, that is, “exact solution for free rotations”.

{\color{blue}{
- same place: ``might become clear'' \textrightarrow ``will become clear'' 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.4, middle of left column: ``for entailing the computation (or numerical estimation) of", replace with ``due to the need to evaluate (or approximate numerically) the". 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- same place: ``a simpler path, though". Not sure it is simpler, so better say ``different path'' and delete ``though". 
}}

\textbf{Answer}:
We did the suggested replacement.

{\color{blue}{
- same place: delete ``implied", since not sure what this word changes in the meaning of ``we attempt to quantify the discretization errors".
}}

\textbf{Answer}:
Done.

{\color{blue}{
- same place: ``average'' \textrightarrow ``averages'' 
}}

\textbf{Answer}:
Done.
{\color{blue}{
- Start of subsection C: ``our proposal'' \textrightarrow ``our approach'' 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.4, after Eq. (10): ``but keep being independent'' \textrightarrow ``but, like U, independent". 
 }}

\textbf{Answer}:
Done.
{\color{blue}{
- p.4, after Eq. (11): ``stacked vector [p pi]". I think this should be [r, q]. Also, not sure why it is necessary to say ``stacked vector", maybe just enough to say ``with respect to [r q]'' 
}}

\textbf{Answer}:
The reviewer is right. We made the correction and removed the expression “stacked vector”.

{\color{blue}{
- p.4, bottom of right column: introduced ``n-th order estimator", but variable n was used earlier to denote sub-steps in the splitting after Eq. (9) and in the middle of right column on p.5. Suggest to use a different letter to avoid confusion. 
}}

\textbf{Answer}:
We thank the reviewer for pointing this out. We now use “k” instead of “n”. We have also replaced “k” by “j” as the index of a few summation signs in the Appendix.

{\color{blue}{
- p.5, middle of right column: ``By this means'' \textrightarrow ``Thus'' 
}}

\textbf{Answer}:
Done.
{\color{blue}{
- the next paragraph: ``intact'' \textrightarrow ``unchanged'' 
}}

\textbf{Answer}:
Done.
{\color{blue}{
- p.6, middle of right column: Change ``poses the question whether'' to ``introduces the question of whether'' and delete ``or not". 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- the next paragraph: what do you mean by ``the angular coefficient"? Is this the slope of the regression line? If yes, then use this term instead of the angular coefficient. 
}}

\textbf{Answer}:
We replaced the original sentence by “It is defined as the slope of a least-square regression line expressing such quantity as a function of time.”

{\color{blue}{
- p.7, middle of left column: ``Leimkhuler'' \textrightarrow ``Leimkuhler".
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.8, middle of left column: ``which differs from the others for involving a global stochastic thermostat", should be ``by involving". 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.8, right column, paragraph starting with ``Fig. 5(d) contains...". The conclusion in this paragraph seems dubious. The authors appear to confuse numerical artefacts, due to finite time step, with the physical phenomenon of heat capacity dependence on temperature. A much more plausible explanation is that statistical error in the measurement of $c_v$ (which is related to the variance of the total energy) is much larger (for the same simulation sample) than that in other quantities (like kinetic or potential energy). So, the statistical error masks the systematic error due to finite time step. In other words, the dependence of measured $c_v$ on time step size is present, but would require more simulation time in order to measure with sufficient precision. 
}}

\textbf{Answer}:
Our (unfulfilled) intention here was to suggest that the effect caused by the numerical artifact is similar to that of a small variation in temperature, in which the average $\langle H \rangle$ changes while $c_v$ remains practically unaltered. We have rewritten the paragraph so as to make this intention clearer. We also employed the expression “are not noticeably influenced” in place of “does not get affected”, in order to acknowledge the statistical nature of these measurements.

{\color{blue}{
- Fig. 5 end of caption: replace ``consists in'' with ``is", as in ``Finally, the KLN scheme (triangle) is the double-thermostat method...". 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.10, bottom of left column: ``Having overcome such barrier ...'' I'm not sure what ``barrier'' the authors have in mind. What they propose allows to increase the accuracy of the results (maybe double it, if we are talking about increasing the time step from 1fs to 4fs), or get the same accuracy with 25% of the computational resource (again from 1fs to 4fs), but this is not overcoming a ``barrier". So, I suggest either deleting this sentence as unnecessary, or modifying it to reflect more accurately the contribution of this work. 
}}

\textbf{Answer}:

\section{\textbf{Reviewer \#2}}

{\color{blue}{
The manuscript presents results from a computational study examining the influence of integration errors on kinetic energy equipartitioning in canonical ensemble MD simulations. The authors argue that an auxiliary shadow kinetic energy should be used instead of the conventional definition when formulating thermostats and analyzing equipartitioning behavior. They show that use of this shadow kinetic energy leads to improved numerical stability and subsequently demonstrate that accurate estimates of properties in the target ensemble can be obtained by reweighting. This reformulation approach is shown to work successfully for several widely-used thermostats.

This is a nicely presented article that is likely to be of interest the readership of JCP. Overall, it is technically sound, but I have a couple of comments that the authors should consider. 
}}

\textbf{Answer}:

{\color{blue}{
(1) There is a recent paper by Smit and co-workers on equipartitioning in thermostatted MD simulations [Smit and co-workers J. Chem. Theory Comput., 2018, 14 (10), pp 5262-5272]. There is no direct overlap with the work presented here, but the authors may wish to add this reference for interested readers and discuss the results in the context of the current study. 
}}

\textbf{Answer}:

{\color{blue}{
(2) The authors assess the magnitude of sampling errors by examining the dependence of average thermodynamics properties on time step. This assumes that correct canonical sampling is recovered when using a small time step, but this may not be true in practice for a number of reasons (e.g., errors in thermostat implementation). The authors should consider analyzing the sample distributions rather than just the averages. To this end, they may wish to consider implementing some of the numerical tests described in J. Chem. Theory Comput., 2013, 9 (2), pp 909-926. 
}}

\textbf{Answer}:

%\bibliographystyle{ieeetr}
%\bibliography{rigid_bodies.bib}
\end{document}

### 2019-02-25 rigid_bodies.tex
Symbolic links for EPS files in main directory
Version submitted to arXiv
add comments on check ensemble results

\usepackage[hidelinks,colorlinks,citecolor=blue]{hyperref}
\usepackage{natbib}
\usepackage{natmove}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{float}
\usepackage{tikz}
\usetikzlibrary{decorations.text}

@@

\begin{abstract}
Kinetic energy equipartition is a premise for many deterministic and stochastic molecular dynamics methods that aim at sampling a canonical ensemble.
While this is expected for real systems, discretization errors introduced by the numerical integration may distort such assumption.
Fortunately, backward error analysis allows us to identify the quantity that is actually subject to equipartition.
This is related to a shadow Hamiltonian, which coincides with the specified Hamiltonian only when the time-step size approaches zero.
This paper deals with discretization effects in a straightforward way.
With a small computational overhead, we obtain refined versions of the kinetic and potential energies, whose sum is a suitable estimator of the shadow Hamiltonian.
Then, we tune the thermostatting procedure by employing the refined kinetic energy instead of the conventional one.
This procedure is shown to reproduce a canonical ensemble compatible with the refined system, as opposed to the original one, but canonical averages regarding the latter can be easily recovered by reweighting.
Water, modeled as a rigid body, is an excellent test case for our proposal because its numerical stability extends up to time steps large enough to yield pronounced discretization errors in Verlet-type integrators.
By applying our new approach, we were able to mitigate discretization effects in equilibrium properties of liquid water for time-step sizes up to 5 fs.
\end{abstract}

@@

For a symplectic integration of Hamiltonian equations of motion, the breakage of equipartition is arguably only apparent \cite{Eastwood_2010}.
The produced trajectory actually follows the dynamics dictated by a \textit{shadow Hamiltonian} \cite{Tuckerman_2010}, which is no longer separable into exclusively momentum- and position-dependent parts and only coincides with the specified Hamiltonian as the time-step size approaches zero.
Therefore, what seems to occur, in fact, is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
As Eastwood \textit{et al}. \cite{Eastwood_2010} have shown, Tolman's generalized equipartition theorem \cite{Tolman_1918} holds at least for small steps, when such a \textit{shadow kinetic energy} keeps displaying a quadratic dependency on the momenta.
The authors then employed backward error analysis to successfully eliminate the artifacts mentioned above.
As we manage to prove here, their findings can be extended to the symplectic integration of rigid-body motion as well.

@@

The new method is specially appealing for rigid-body MD, primarily because numerical stability can, in this case, extend up to time-step sizes large enough to yield pronounced discretization errors.
For rigid water, which served as an excellent test case, we were able to correct discretization errors in ensemble averages computed with time-step sizes up to $5~\text{fs}$, thus outperforming not only the KLN method \cite{Kamberaj_2005}, but also a reimplementation for rigid bodies of the NHC integrator introduced by Martyna \textit{et al}. \cite{Martyna_1996}, as well as stochastic rescaling \cite{Bussi_2007} applied simultaneously to translational and angular velocities.

@@

In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the quaternion-conjugate momentum.
Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, and $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
Matrices $\mt M$ and $\mt I$ are diagonal ones.
The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
Finally, the matrix-valued functions $\mt B$ and $\vt C$, given by
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \left[\begin{array}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{array}\right]
\end{equation*}
and
\begin{equation*}
\mt C(\vt q) = \left[\begin{array}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{array}\right]
\end{equation*}
are related to the rotation matrix ${\mt A}(\vt q)$ through the factorization ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$ that was mentioned at the beginning of this section. Eq.~\eqref{eq:ODE system for NVE} preserves the constraints $\tr{\vt q}{\vt q} = 1$ and $\tr{\vt q}{\vt \pi} = 0$, as well as the value of a Hamiltonian
\begin{equation}
\label{eq:Hamiltonian}
\Ham{} = K(\vt p, \vt q, \vt \pi) + U(\vt r,\vt q),
\end{equation}
where $K$ and $U$ are the kinetic and potential energies of the body, respectively.
While $U(\vt r, \vt q)$ is an arbitrary function, the kinetic energy is given by
\begin{equation}
\label{eq:kinetic energy}
K = \frac{1}{2} \tr{\vt p} {\mt M}^{-1} {\vt p} + \frac{1}{8} \tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}.
\end{equation}

Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the corresponding properties of all bodies piled together.
The size of diagonal matrices $\mt M$ and $\mt I$ also increases to $3N \times 3N$ in this case.
In addition, $\mt B(\vt q)$ and $\mt C(\vt q)$ become block-diagonal matrices with $4N$ rows and $3N$ columns.
Yet, the notation can be made even more general if $\vt r$, $\vt p$, and $\mt M$ are enlarged further so as to accommodate a number of freely moving atoms.

@@

A symplectic, time-reversible integrator can be devised by splitting the exponential operator according to the Trotter-Suzuki formula \cite{Trotter_1959, Suzuki_1976}
\begin{equation}
\label{eq:splitting NVE}
e^{\timestep \Liu{\tiny NVE}} = e^{\frac{\timestep}{2} \Liu{B}} e^{\timestep \Liu{A}} e^{\frac{\timestep}{2} \Liu{B}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$.
In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively; where the asterisk denotes the state of the system immediately before the propagation occurs.
The action of $e^{\timestep \Liu{A}}$ is a simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation along a full time step.
We have derived \cite{Silveira_2017} an exact solution for torque-free rotations in a form that facilitates computer implementation and allows, differently from other existing solutions \cite{Kosenko_1998, vanZon2007, Celledoni_2008}, a unified treatment of asymmetric, symmetric, and spherical tops.
This solution was used to evaluate the simpler NO-SQUISH method \cite{Miller_2002}.
Despite its approximate treatment of rotations, which are split into several revolutions around the principal axes \cite{Dullweber_1997}, that method proved to be very accurate in liquid-phase simulations \cite{Silveira_2017}.
This justifies its more widespread use.
However, having an unplit solution for rotations at hand is crucial for the feasibility of backward error analysis, as it might become clear in Sec.~\ref{sec:backward error analysis} and Appendix \ref{sec:rigid body shadow hamiltonian}.

@@

Due to its low computational cost, the whole procedure described above can be repeated $n$ times with little impact on the overall integration effort.
This makes it feasible to increase the time step $\timestep$ without ruining the accuracy of the NHC part.
In contrast, evaluating the NVE part is expensive for involving force/torque computations and, in addition, its accuracy goes down quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
A high-order integration scheme\cite{Omelyan_2007, Van_zon_2008} could possibly admit larger time steps, but it would raise both cost and complexity for entailing the computation (or numerical estimation) of force/torque gradients.
Omelyan's processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
Here we choose to take a simpler path, though.
Instead of trying to increase accuracy in the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we attempt to quantify the implied discretization errors, with which we can both tune the thermostatting procedure and properly weight the sampled configurations when ensemble average are computed.

@@

In order to explain our proposal, we turn the attention again to the NVE case.
A known property of splitting methods applied to Hamiltonian ODE systems is that they provide approximate solutions which are Hamiltonian as well.
Hence, a trajectory generated to be roughly consistent with the specified function $\Ham{}$ will be exactly consistent with a nearby (albeit unknown) function $\Ham{S}$, referred to as a shadow Hamiltonian \cite{Hairer_2006}, which explicitly depends on the time-step size $\timestep$.

@@

Note that $\refined U$ becomes dependent on $\timestep^2$, but keeps being independent of $\vt p$ and $\vt \pi$.
In turn, the \textit{refined kinetic energy} can be expressed as
\begin{equation}
\label{eq:modified kinetic energy}
\refined K = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
where $\refined{\mathbf \Omega}$ is a matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$ given by $\tilde{\mt \Omega} = {\mt \Omega} + ({\timestep^2}/{6}) {\mt \Omega} {\mt \Xi} \tr{\mt \Omega}$.
In this definition, $\mt \Xi$ is the Hessian matrix of the potential energy with respect to the stacked vector $[\substack{\vt p \\ \vt \pi}]$, while ${\mt \Omega}$ is a block-diagonal matrix defined so that $K = \frac{1}{2} \tr{[\substack{\vt p \\ \vt \pi}]} {\mt \Omega} [\substack{\vt p \\ \vt \pi}]$.
Although $\refined{\mathbf \Omega}$ is symmetric like $\mt \Omega$, its structure is not block-diagonal due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both $\vt r$ and $\vt q$.
As a result, one cannot generally split the refined kinetic energy into independent translational and rotational contributions.

@@

Analytical calculation of $\refined U$ via Eq.~\eqref{eq:modified potential energy} is straightforward.
In contrast, although it is possible to compute $\refined K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
Fortunately, we can estimate it rather easily by following Eastwood \textit{et al}. \cite{Eastwood_2010}, who employed numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
\begin{equation}
\label{eq:modified kinetic energy estimator}
\refined K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(\timestep^n).
\end{equation}

>>>>

\usepackage[hyperindex,breaklinks,hidelinks,colorlinks,citecolor=blue]{hyperref}
\usepackage{natbib}
\usepackage{natmove}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{float}
\usepackage{tikz}
\usetikzlibrary{decorations.text}

@@

\begin{abstract}
Kinetic energy equipartition is a premise for many deterministic and stochastic molecular dynamics methods that aim at sampling a canonical ensemble.
While this is expected for real systems, discretization errors introduced by the numerical integration may lead to deviations from equipartition.
Fortunately, backward error analysis allows us to obtain a higher-order estimate of the quantity that is actually subject to equipartition.
This is related to a shadow Hamiltonian, which converges to the specified Hamiltonian only when the time-step size approaches zero.
This paper deals with discretization effects in a straightforward way.
With a small computational overhead, we obtain refined versions of the kinetic and potential energies, whose sum is a suitable estimator of the shadow Hamiltonian.
Then, we tune the thermostatting procedure by employing the refined kinetic energy instead of the conventional one.
This procedure is shown to reproduce a canonical ensemble compatible with the refined system, as opposed to the original one, but canonical averages regarding the latter can be easily recovered by reweighting.
Water, modeled as a rigid body, is an excellent test case for our proposal because its numerical stability extends up to time steps large enough to yield pronounced discretization errors in Verlet-type integrators.
By applying our new approach, we were able to mitigate discretization effects in equilibrium properties of liquid water for time-step sizes up to 5 fs.
\end{abstract}

@@

For a symplectic integration of Hamiltonian equations of motion, the breakdown of equipartition is arguably only apparent \cite{Eastwood_2010}.
The produced trajectory actually follows the dynamics dictated by a \textit{shadow Hamiltonian} \cite{Tuckerman_2010}, which is no longer separable into exclusively momentum- and position-dependent parts and only coincides with the specified Hamiltonian as the time-step size approaches zero.
Therefore, what seems to occur, in fact, is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
As Eastwood \textit{et al}. \cite{Eastwood_2010} have shown, Tolman's generalized equipartition theorem \cite{Tolman_1918} holds at least for small steps, when such a \textit{shadow kinetic energy} keeps displaying a quadratic dependency on the momenta.
The authors then employed backward error analysis to successfully eliminate the artifacts mentioned above.
As we manage to demonstrate here, their findings can be extended to the symplectic integration of rigid-body motion as well.

@@

The new method is especially appealing for rigid-body MD, primarily because numerical stability can, in this case, extend up to time-step sizes large enough to yield pronounced discretization errors.
For rigid water, which served as an excellent test case, we were able to correct discretization errors in ensemble averages computed with time-step sizes up to $5~\text{fs}$, thus outperforming not only the KLN method \cite{Kamberaj_2005}, but also a reimplementation for rigid bodies of the NHC integrator introduced by Martyna \textit{et al}. \cite{Martyna_1996}, as well as stochastic rescaling \cite{Bussi_2007} applied simultaneously to translational and angular velocities.

@@

In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the quaternion-conjugate momentum.
Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, and $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
Matrices $\mt M$ and $\mt I$ are diagonal ones.
The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
Finally, the matrix-valued functions $\mt B$ and $\vt C$, given by
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \left[\begin{array}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{array}\right]
\end{equation*}
and
\begin{equation*}
\mt C(\vt q) = \left[\begin{array}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{array}\right]
\end{equation*}
are related to the rotation matrix ${\mt A}(\vt q)$ through the factorization ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$ that was mentioned at the beginning of this section. Eq.~\eqref{eq:ODE system for NVE} preserves the constraints $\tr{\vt q}{\vt q} = 1$ and $\tr{\vt q}{\vt \pi} = 0$, as well as the value of a Hamiltonian
\begin{equation}
\label{eq:Hamiltonian}
\Ham{} = K(\vt p, \vt q, \vt \pi) + U(\vt r,\vt q),
\end{equation}
where $K$ and $U$ are the kinetic and potential energies of the body, respectively.
While the form of $U(\vt r, \vt q)$ depends on the specific interaction model, the kinetic energy is given by
\begin{equation}
\label{eq:kinetic energy}
K = \frac{1}{2} \tr{\vt p} {\mt M}^{-1} {\vt p} + \frac{1}{8} \tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}.
\end{equation}

Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the corresponding properties of all bodies combined in a single vector.
The size of diagonal matrices $\mt M$ and $\mt I$ also increases to $3N \times 3N$ in this case.
In addition, $\mt B(\vt q)$ and $\mt C(\vt q)$ become block-diagonal matrices with $4N$ rows and $3N$ columns.
Yet, the notation can be made even more general if $\vt r$, $\vt p$, and $\mt M$ are enlarged further so as to accommodate a number of individual point masses.

@@

A symplectic, time-reversible integrator can be devised by splitting the exponential operator according to the Trotter-Suzuki formula \cite{Trotter_1959, Suzuki_1976}
\begin{equation}
\label{eq:splitting NVE}
e^{\timestep \Liu{\tiny NVE}} = e^{\frac{\timestep}{2} \Liu{B}} e^{\timestep \Liu{A}} e^{\frac{\timestep}{2} \Liu{B}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$.
In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick that changes the linear and quaternion momenta according to ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively; where the asterisk denotes the state of the system immediately before the propagation occurs.
The action of $e^{\timestep \Liu{A}}$ is a simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation along a full time step.
We have derived \cite{Silveira_2017} an exact solution for torque-free rotations in a form that facilitates computer implementation and allows, differently from other existing solutions \cite{Kosenko_1998, vanZon2007, Celledoni_2008}, a unified treatment of asymmetric, symmetric, and spherical tops.
This solution was used to evaluate the simpler NO-SQUISH method \cite{Miller_2002}.
Despite its approximate treatment of rotations, which are split into several revolutions around the principal axes \cite{Dullweber_1997}, that method proved to be very accurate in liquid-phase simulations \cite{Silveira_2017}.
This justifies its more widespread use.
However, having an exact solution for free rotations at hand is crucial for the feasibility of backward error analysis, as it will become clear in Sec.~\ref{sec:backward error analysis} and Appendix \ref{sec:rigid body shadow hamiltonian}.

@@

Due to its low computational cost, the whole procedure described above can be repeated $n$ times with little impact on the overall integration effort.
This makes it feasible to increase the time step $\timestep$ without ruining the accuracy of the NHC part.
In contrast, evaluating the NVE part is expensive for involving force/torque computations and, in addition, its accuracy goes down quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
A high-order integration scheme\cite{Omelyan_2007, Van_zon_2008} could possibly admit larger time steps, but it would raise both cost and complexity due to the need to evaluate (or approximate numerically) the force/torque gradients.
Omelyan's processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
Here we choose to take a different path.
Instead of trying to increase accuracy in the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we attempt to quantify the discretization errors, with which we can both tune the thermostatting procedure and properly weight the sampled configurations when ensemble averages are computed.

@@

In order to explain our approach, we turn the attention again to the NVE case.
A known property of splitting methods applied to Hamiltonian ODE systems is that they provide approximate solutions which are Hamiltonian as well.
Hence, a trajectory generated to be roughly consistent with the specified function $\Ham{}$ will be exactly consistent with a nearby (albeit unknown) function $\Ham{S}$, referred to as a shadow Hamiltonian \cite{Hairer_2006}, which explicitly depends on the time-step size $\timestep$.

@@

Note that $\refined U$ becomes dependent on $\timestep^2$ but, like $U$, it is independent of $\vt p$ and $\vt \pi$.
In turn, the \textit{refined kinetic energy} can be expressed as
\begin{equation}
\label{eq:modified kinetic energy}
\refined K = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
where $\refined{\mathbf \Omega}$ is a matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$ given by $\tilde{\mt \Omega} = {\mt \Omega} + ({\timestep^2}/{6}) {\mt \Omega} {\mt \Xi} \tr{\mt \Omega}$.
In this definition, $\mt \Xi$ is the Hessian matrix of the potential energy with respect to $[\substack{\vt r \\ \vt q}]$, while ${\mt \Omega}$ is a block-diagonal matrix defined so that $K = \frac{1}{2} \tr{[\substack{\vt p \\ \vt \pi}]} {\mt \Omega} [\substack{\vt p \\ \vt \pi}]$.
Although $\refined{\mathbf \Omega}$ is symmetric like $\mt \Omega$, its structure is not block-diagonal due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both $\vt r$ and $\vt q$.
As a result, one cannot generally split the refined kinetic energy into independent translational and rotational contributions.

@@

Analytical calculation of $\refined U$ via Eq.~\eqref{eq:modified potential energy} is straightforward.
In contrast, although it is possible to compute $\refined K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
Fortunately, we can estimate it rather easily by following Eastwood \textit{et al}. \cite{Eastwood_2010}, who employed numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
Considering $k$-th order estimators $\dot{\vt r}^{[k]}$ and $\dot{\vt q}^{[k]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
\begin{equation}
\label{eq:modified kinetic energy estimator}
\refined K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[k]} + \tr{\vt \pi} \dot{\vt q}^{[k]} \right) + \mathcal{O}(\timestep^k).
\end{equation}
